Separating random and deterministic sources of computational noise in explore-exploit decisions

### 2024-05-16 manuscript/2024 - Plot Comp Bio/manuscript_DeterministicRandomNoise.tex
noise paper

\date{\today}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 15$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 85$\%$ of the variability is truly `random,' although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by deterministic processing of the stimulus (which we refer to as `deterministic noise') vs other processes, including deterministic processing of unobserved stimuli as well as truly stochastic processes (which we refer to as `random noise'). To distinguish between these two types of noise, we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus (i.e., decision noise is purely deterministic noise), then people's choices should be identical for both decisions, since the stimulus is the same both times. Conversely, if the decision is a purely random function of the stimulus (i.e., decision noise is purely random noise), then people's choices will be different 50$\%$ of the time, since the random noise is different each time. In between these two extremes of purely deterministic and purely random drivers of behavioral variability, the extent to which people's decisions are consistent between the two decisions can be used to estimate the amount of deterministic and random noise.

@@

	The task is organized in games (Figure \ref{fig:taskfig}A). The mean payout of the two bandits are held fixed within a game and reset between games. Each game consists of either 5 or 10 trials. In the first 4 trials, participants are instructed about which bandit to choose, this allows us to manipulate what information from both bandits participants receive before they make their first free choice between the two bandits. From the 5th trial, participants make free choices between the two bandits. Participants have either 1 or 6 free choices to make.

@@

	The second key feature of the Horizon Task are the 4 `forced-choice trials' at the start of each game that allow us to control exactly what participants know about the two bandits before they make their choice. In these forced-choice trials, participants are instructed which of the bandits to play allowing us to control how much information they have about each of the options. The forced-choice trials are used to set up one of two information conditions: an `unequal information' or [1 3] condition, in which participants play one bandit once and the other three times, and an `equal information' or [2 2] condition, in which participants play both bandits twice.

	Relative to the original Horizon Task, the key modification in this paper is to give people `repeated games' (Figure \ref{fig:taskfig}B), in which they see exact same set of forced-choice plays twice in two separate games separated by several minutes in time so as to avoid detection. By repeating the forced-choice plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, then they will make the same decision in both games and their choices will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games will be inconsistent some fraction of the time. The extent to which participants' choices are consistent on the repeated versions of the games allow us to quantify the extent to which the variability in their behavior was driven by a deterministic process vs a random noise process.

@@

	\subsubsection*{Model-free analysis}
	In the model-free analysis, we quantify random and directed exploration using simple choice probabilities. Random exploration is quantified as the probability of choosing the option that has the lower average payout in the forced-choice plays in the equal, or [2 2], condition, $p(\mbox{low mean})$. The idea here is that, in the equal condition, the optimal strategy is to compute the mean payout for each bandit from the example plays and then always choose the option with the highest mean. When participants do not choose the option with the higher mean, the assumption is that this is due to some kind of `decision noise', making the probability of choosing the low mean option a measure of behavioral variability. In this view, random exploration corresponds to an increase in $p(\mbox{low mean})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}A; t(64) = 7.99, p $<$ 0.001).

@@

	The free parameters of this model are: the information bonus $A$, which controls the level of directed exploration; the noise standard deviation, $\sigma$, which controls the level of random exploration, and the spatial bias, $b$, which determines the extent to which participants prefer the option on the right. These free parameters are fit separately for each participant in each horizon and information condition, allowing us to test whether directed and random exploration increase with horizon. Consistent with previous research, we find that this is indeed the case (Figure \ref{fig:modelfree}C; t(64) = 5.35, p $<$ 0.001. Figure \ref{fig:modelfree}D; t(64) = 3.54, p $<$ 0.001).

@@

We used hierarchical Bayesian analysis to fit the parameters of the model (see Figure \ref{fig:model} for a graphical representation of the model in the style of \cite{lee2014}). In particular, we fit values of the information bonus $A$, spatial bias $b$, variance of random noise $\sigma_{ran}^2$, and variance of deterministic noise, $\sigma_{det}^2$ for each participant in each horizon. Model fitting was performed using the MATJAGS and JAGS software \citep{jags, matjags} with full details given in the Methods.

\subsubsection*{Model validation \label{ch:appendix:bayesrecovery}}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters. Full details of this analysis are presented in the Supplementary Materials.

@@

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (Supplementary Figure XXX). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions.
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

Thus, overall, we were able to detect both deterministic and random noises using our model. Our model provides a lower bound for deterministic noise and an upper bound for random noise. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games.

@@

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior.jpg}
	\caption{Model based analysis showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.}
	%BOB SAYS: DON'T NEED THE NUMBERS ON THE Y-AXES. 
	\label{fig:mb1}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.5\textwidth]{figures/RDBayes_ratio__paironly.jpg}
	\caption{Model based analysis showing the posterior distributions over the ratio of the group-level mean of the standard deviations of  random and deterministic noise between horizon 6 and horizon 1 respectivelly. The ratio in the standard deviations of noise between horizon 6 and horizon 1 is similar for random and deterministic noise.}
	\label{fig:ratio}
\end{center}
\end{figure}

@@

To perform this `posterior predictive check,' we created a set of simulated data by taking the subject-level parameters from the hierarchical Bayesian fits and having the model play the same sequence of games as seen by the subjects. We then applied the same model-free analysis as described in the previous sections to this simulated data set and compared the model's behavior to that of participants. As shown in Figure \ref{fig:mb3}, the model can account for all qualitative patterns in the data --- the increase in p(high info), p(low mean), and p(inconsistent) with horizon, and that p(inconsistent) is in between pure random and pure deterministic noise. The quantitative agreement is almost perfect for p(high info) and for p(inconsistent) in the [1 3] condition, but the model seems to systematically overestimate p(low mean) and p(inconsistent) in [2 2] conditions, although the discrepancy is relatively small (overestimating p(low mean) by 0.054 or 31.37\%, and p(inconsistent) by 0.049 or 27.83\% in [2 2] condition).

\begin{figure}[H]
\begin{center}
	\includegraphics[width=1\textwidth]{figures/RDBayes_2noise_modelA.jpg}
	\caption{
		Our model accounts for all qualitative patterns of the data, namely, p(high info) and p(low mean) increase as a function of horizon, p(inconsistent) increases as a function of horizon for both [1 3] and [2 2] conditions and lies between the pure random and pure deterministic noise prediction.}
	\label{fig:mb3}
\end{center}
\end{figure}

@@

\section*{Discussion}
In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise.'  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise.' In particular, we found that deterministic noise accounted for around 15\% of the overall variability in people's behavior.

@@

One interpretation for this low level of deterministic noise is that most of the variability in random exploration is truly random. Such a random noise interpretation, would be consistent with recent work showing that variability in perceptual decisions may be driven by imperfections in mental inference \cite{drugowitsch16}. In this view, apparently random behavior is not due to sensory processing or response selection, but to suboptimal computations in the brain. Although suboptimal inference is different from simply adding random noise to neural circuitry\citep{Pouget12}, as long as the suboptimality in neural computation is not a deterministic function of the stimuli, it is a form of random noise in our definition. Indeed, a strong interpretation of this hypothesis would suggest that randomness in explore-exploit behavior is due to imperfect inference about the correct course of action. In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute \citep{Wilson2020}.

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 85\% of the decision noise that is not stimulus-driven noise, could be deterministic.

@@

Such a reduction in the strength of reward coding in exploration, is consistent with our recent work using a drift diffusion model (DDM) to model explore-exploit decisions \citep{Feng21}. In the drift diffusion model, changes in behavioral variability can be driven by changes in the decision threshold (smaller threshold = more noise) or changes in the signal-to-noise ratio with which reward is encoded (lower SNR = more noise). By fitting both choices and response times, we were able to distinguish between these two accounts showing that the majority of the horizon-change in variability was driven by changes in SNR not threshold. However, this model could not determine whether the changes in SNR were driven by signal or noise. By showing that the change in deterministic and random noise have the same ratio, the present work suggests that this SNR change is driven by changes in reward-signal processing, not noise. Of course, to truly see whether changes in signal or noise are driving random exploration will require more direct measurements of neural processing such as with neuroimaging and electrophysiology \citep{Tomov20, Ebitz18,costa1,costa22}

@@

80 participants (ages 18-25, 37 male, 43 female) from the University of Arizona undergraduate subject pool participated in the experiment. 14 were excluded on the basis of performance, using the same exclusion criterion as in \citep{wilson2014}. In this exclusion criteria, we measured the accuracy of each participant's choices by calculating the percentage of times that a participant chose the bandit with the higher underlying mean payouts in the last choice of a long horizon game, intuitively people should figure out which bandit has a higher mean payout by the last trial and should have an accuracy measure significantly above 50\%, specifically, we computed the likelihood that the measured accuracy can be achieved by making a completely random choice between the two options and excluded participants with a likelihood smaller than 99.999\%, in other words, participants who didn't show an accuracy significant above chance with $p < 0.001$ were excluded in the analysis. This left 65 for the main analysis. Note that including the 15 badly performing subjects did not change the main results (Supplementary Figures S1, S2, S11)

@@

Each game consisted of 5 or 10 choices. Every game started with a fixation cross, then a bar of boxes appeared indicating the horizon for that game. For the first 4 trials - the instructed trials, we highlight the box on one of the bandits to instruct the participant to choose that option. On these trials, they have to press the corresponding key to reveal the outcome. From the fifth trial, boxes on both bandits will be highlighted and they are free to make their own decision. There was no time limit for decisions. During free choices participants could press either the left arrow key or right arrow key to indicate their choice of left or right bandit. The score feedback was presented for 300ms. The task was programmed using Psychtoolbox in MATLAB \citep{psychtoolbox1, psychtoolbox2}.

The first four trials of each game were forced-choice trials, in which only one of the options was available for the participant to choose. We used these forced-choice instructed trials to manipulate the relative ambiguity of the two options, by providing the participant with different amounts of information about each bandit before their first free choice. The four forced-choice trials set up two uncertainty conditions: unequal uncertainty(or [1 3]) in which one option was forced to be played once and the other three times, and equal uncertainty(or [2 2]) in which each option was forced to be played twice. After the forced-choice trials, participants made either 1 or 6 free choices (two horizon conditions),  Figure \ref{fig:taskfig}.

\subsection*{Model-based analysis}
We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model in \citep{wilson2014} that was modified to differentiate deterministic noise from random noise. Because the stimuli are identical in the repeated games, by definition, deterministic noise remains the same in repeated games, whereas random noise can change.

>>>>

%\date{\today}

@@

	\newpage
	\begin{abstract}
		Human decision making is inherently variable. While this variability is often seen as a sign of suboptimal behavior, recent work suggests that variability can actually be adaptive. An example arises when we must choose between exploring unknown options or exploiting options we know well. A little randomness in these `explore-exploit' decisions is remarkably effective as it can encourage us to explore options we might otherwise ignore. In line with this idea, several studies have found evidence that people increase their behavioral variability when it is valuable to explore. A key question, however, is whether this variability in so-called `random exploration' is actually random. That is, is random exploration driven by stochastic processes in the brain or by some unobserved deterministic process that we have failed to account for when measuring behavioral variability? By designing an explore-exploit task in which, unbeknownst to them, participants are presented with the exact same choice twice, we provide a partial answer to this question. By modeling behavior in this task, we were able to estimate a lower bound on the amount of variability that is deterministically driven by the stimulus and an upper bound on the amount of variability that is random. Using this approach, we find evidence that at least 14$\%$ of the variability in random exploration in our studied task can be accounted for by deterministic processing of the stimulus. Conversely, this suggests that up to 86$\%$ of the variability is truly `random,' although it is still possible that this variability is driven by deterministic factors not related to the stimulus. Finally, our results suggest that both deterministic and random sources of variability change proportionally to each other as the value of exploration increases, suggesting that a common noise gating mechanism may be at play in random exploration.		
	\end{abstract}
	\newpage
	\section*{Introduction}
	%BOB SAID: IN GENERAL, NEED TO GO THROUGH THE PAPER AND MAKE SURE THAT `RANDOM' NOISE IS REPLACES WITH `NOT A DETERMINISTIC FUNCTION OF THE STIMULUS' OR SOMETHING LIKE THAT

@@

	In this paper, we investigate the extent to which the apparent randomness in random exploration can be explained by deterministic processing of the stimulus (which we refer to as `deterministic noise') vs other processes, including deterministic processing that is unrelated to the stimuli as well as truly stochastic processes (which we refer to as `random noise'). To distinguish between these two types of noise, we modify the Horizon Task \citep{wilson2014} to have people face the exact same explore-exploit choice twice. If the decision is a purely deterministic function of the stimulus (i.e., decision noise is purely deterministic noise), then people's choices should be identical for both decisions, since the stimulus is the same both times. Conversely, if the decision is a purely random function of the stimulus (i.e., decision noise is purely random noise), then people's choices will be different 50$\%$ of the time, since the random noise is different each time. In between these two extremes of purely deterministic and purely random drivers of behavioral variability, the extent to which people's decisions are consistent between the two decisions can be used to estimate the amount of deterministic and random noise.

@@

	The task is organized in games (Figure \ref{fig:taskfig}A). The mean payout of the two bandits are held fixed within a game and reset between games. Each game consists of either 5 or 10 trials. The first four trials of each game are `forced-choice' trials. In the first four trials, participants are instructed about which bandit to choose, this allows us to manipulate what information from both bandits participants receive before they make their first free choice between the two bandits. From the 5th trial, participants make free choices between the two bandits. Participants have either 1 or 6 free choices to make.

@@

	The second key feature of the Horizon Task are the 4 forced-choice trials at the start of each game that allow us to control exactly what participants know about the two bandits before they make their choice. In these forced-choice trials, participants are instructed which of the bandits to play allowing us to control how much information they have about each of the options. The forced-choice trials are used to set up one of two information conditions: an `unequal information' or [1 3] condition, in which participants play one bandit once and the other three times, and an `equal information' or [2 2] condition, in which participants play both bandits twice.

	Relative to the original Horizon Task, the key modification in this paper is to give people `repeated games' (Figure \ref{fig:taskfig}B), in which they see the exact same set of forced-choice plays twice in two separate games separated by several minutes in time so as to avoid detection. By repeating the forced-choice plays for each game twice, we can set up a situation where (unbeknownst to the participants) they are faced with the exact same explore-exploit choice, with the exact same stimuli twice. Thus, if their behavior is a deterministic function of the stimuli, then they will make the same decision in both games and their choices will be consistent. Conversely, if their behavior is not driven by a deterministic function of the stimulus, then their choices on the repeated games will be inconsistent some fraction of the time. The extent to which participants' choices are consistent on the repeated versions of the games allow us to quantify the extent to which the variability in their behavior was driven by a deterministic process vs a random noise process.

@@

	\subsubsection*{Model-free analysis}
	In the model-free analysis, we quantify random and directed exploration using simple choice probabilities. Random exploration is quantified as the probability of choosing the option that has the lower average payout in the forced-choice plays in the equal, or [2 2], condition, $p(\mbox{low mean})$. The idea here is that, in the equal condition, the optimal strategy is to compute the mean payout for each bandit from the forced-choice plays and then always choose the option with the highest mean. When participants do not choose the option with the higher mean, the assumption is that this is due to some kind of `decision noise', making the probability of choosing the low mean option a measure of behavioral variability. In this view, random exploration corresponds to an increase in $p(\mbox{low mean})$ with horizon, which is exactly what we see in the data (Figure \ref{fig:modelfree}A; t(64) = 7.99, p $<$ 0.001).

@@

	The free parameters of this model are: the information bonus $A$, which controls the level of directed exploration; the noise standard deviation, $\sigma$, which controls the level of random exploration, and the spatial bias, $b$, which determines the extent to which participants prefer the option on the right. These free parameters are fit separately for each participant in each horizon condition, allowing us to test whether directed and random exploration increase with horizon. Consistent with previous research, we find that this is indeed the case (Figure \ref{fig:modelfree}C; t(64) = 5.35, p $<$ 0.001. Figure \ref{fig:modelfree}D; t(64) = 3.54, p $<$ 0.001).

@@

We used hierarchical Bayesian analysis to fit the parameters of the model (see Figure \ref{fig:model} for a graphical representation of the model in the style of \cite{lee2014}). In particular, we fit values of the information bonus $A$, spatial bias $b$, variance of random noise $\sigma_{ran}^2$, and variance of deterministic noise $\sigma_{det}^2$ for each participant in each horizon. Model fitting was performed using the MATJAGS and JAGS software \citep{jags, matjags} with full details given in the Methods.

\subsubsection*{Model validation}
To be sure that our fit parameter values were meaningful and to understand the limits of our model, we evaluated our model extensively using simulated data. This allowed us to quantify whether deterministic and random noise can be identified under ideal conditions where the behavior is generated by the model with known parameters.

@@

Parameter recover in this task was good for this model, with fit values for $\sigma_{ran}$ and $\sigma_{det}$ showing strong correlations with their simulated values in both horizon (H) conditions (For $\sigma_{ran}$, $R = 0.91 (\mbox{H} = 1) \mbox{ and } 0.84 (\mbox{H} = 6), p < 0.001$, For $\sigma_{det}$, $R = 0.60 (\mbox{H} = 1) \mbox{ and } 0.59 (\mbox{H} = 6), p < 0.001$). However, while the relationship was near perfect for random noise ($\frac{\mbox{\small Recovered } \sigma_{ran}}{\mbox{\small Simulated } \sigma_{ran}} = 1.01$), there was a systematic bias to underestimate the level of deterministic noise by about 32\% ($\frac{\mbox{\small Recovered } \sigma_{det}}{\mbox{\small Simulated } \sigma_{det}} = 0.68$). Despite this underestimation of deterministic noise in both horizon conditions, the difference in deterministic noise between horizons is much better captured (See Supplementary Materials section 2.2). This is because the underestimation of deterministic noise is partially canceled out when the difference is taken between horizon conditions. In addition, we see better parameter recovery for random noise than deterministic noise. This is likely because we effectively have half as many trials for deterministic noise. In particular, while we generate two samples of random noise for each repeated game pair, we only generate one sample of deterministic noise, which by definition is the same in both of the repeated games. 
	% $\frac{\mbox{\small Recovered } \Delta \sigma_{det}}{\mbox{\small Simulated } \Delta \sigma_{det}} = 0.74$

In addition to the conventional subject-level parameter recovery analysis presented here, we also performed parameter recovery analysis that examined how faithful the full posterior distribution of group-level parameters can be recovered in simulated data (Supplementary Figure S5, S6). Qualitatively, we also showed that our way of modeling deterministic noise is capable of capturing known deterministic processes intentionally omitted from the full model (Supplementary Figure S4). Full details of these additional analyses are presented in the Supplementary Materials.

Overall, we were able to detect both deterministic and random noises using our model. Because random noise is modeled as non-stimulus-driven noise, it can reflect both true stochastic random noise and possible deterministic noises which do not depend on the stimuli. Thus conceptually our random noise estimate provides an upper bound for the true `random noise' induced by intrinsic stochastic processes in the brain. Thus, our model provides a lower bound for deterministic noise and an upper bound for random noise.

@@

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.7\textwidth]{figures/RDBayes_hyperprior.jpg}
	\caption{Model based analysis showing the posterior distributions over the group-level mean of the standard deviations of  random and deterministic noise. Both random (A, B) and deterministic (C,D) noises are nonzero (A, C) and change with horizon (B, D).  }
	%BOB SAYS: However, random noise has both a greater magnitude overall (A, C) and a greater change with horizon (B, D) than deterministic noise.
	\label{fig:mb1}
\end{center}
\end{figure}

\begin{figure}[H]
\begin{center}
	\includegraphics[width=0.5\textwidth]{figures/RDBayes_ratio__paironly.jpg}
	\caption{Model based analysis showing the posterior distributions over the ratio of the group-level mean of the standard deviations of  random and deterministic noise between horizon 6 and horizon 1 respectivelly. The ratio in the standard deviations of noises between horizon 6 and horizon 1 is similar for random and deterministic noise.}
	\label{fig:ratio}
\end{center}
\end{figure}

@@

To perform this `posterior predictive check', we created a set of simulated data by taking the subject-level parameters from the hierarchical Bayesian fits and having the model play the same sequence of games as seen by the subjects. We then applied the same model-free analysis as described in the previous sections to this simulated data set and compared the model's behavior to that of participants. As shown in Figure \ref{fig:mb3}, the model can account for all qualitative patterns in the data --- the increase in p(high info), p(low mean), and p(inconsistent) with horizon, and that p(inconsistent) is in between pure random and pure deterministic noise. The quantitative agreement is almost perfect for p(high info) and for p(inconsistent) in the [1 3] condition, but the model seems to systematically overestimate p(low mean) and p(inconsistent) in [2 2] conditions, although the discrepancy is relatively small (overestimating p(low mean) by 0.054 or 31.37\%, and p(inconsistent) by 0.049 or 27.83\% in [2 2] condition).

As a control, we also applied posterior predictive checks on alternative models that consider only deterministic or only random noise, and these models fail to capture all qualitative patterns (Supplementary Figure S13). Full details of this analysis can be found in Supplementary Materials section 3.2.

\begin{figure}[H]
\begin{center}
	\includegraphics[width=1\textwidth]{figures/RDBayes_2noise_modelA.jpg}
	\caption{
		Our model accounts for all qualitative patterns of the data, namely, p(high info) and p(low mean) increase as a function of horizon, p(inconsistent) increases as a function of horizon for both [1 3] and [2 2] conditions and it lies between the pure random and pure deterministic noise prediction.}
	\label{fig:mb3}
\end{center}
\end{figure}

@@

\section*{Discussion}
In this paper, we investigated whether random exploration is really random or whether it is driven deterministically by aspects of the stimulus we have previously ignored when measuring `decision noise.'  Using a version of the Horizon Task with repeated games, we found evidence that at least some of the noise in random exploration could be explained by such `deterministic noise.' In particular, we found that deterministic noise accounted for around 14\% of the overall variability in people's behavior.

@@

One interpretation for this low level of deterministic noise is that most of the variability in random exploration is truly random. Such a random noise interpretation, would be consistent with recent work showing that variability in perceptual decisions may be driven by imperfections in mental inference \citep{drugowitsch16}. In this view, apparently random behavior is not due to sensory processing or response selection, but to suboptimal computations in the brain. Although suboptimal inference is different from simply adding random noise to neural circuitry\citep{Pouget12}, as long as the suboptimality in neural computation is not a deterministic function of the stimuli, it is a form of random noise in our definition. Indeed, a strong interpretation of this hypothesis would suggest that randomness in explore-exploit behavior is due to imperfect inference about the correct course of action. In the context of the Horizon Task, such computational errors would likely be larger in the long horizon condition as the correct course of action in these cases is much harder to compute \citep{Wilson2020}.

Although the random noise interpretation is theoretically appealing, our approach, while an improvement on previous methods, is not without limitations. Most important is that our measure of `random' noise is only an upper bound on the true level of randomness and that, in principle, the random decision noise could be lower. Specifically, in our model, what we labeled random noise was really `non-stimulus-driven variability'. While this non-stimulus-driven variability could be driven by truly random stochastic processes, it could also be driven by deterministic processing that is unrelated to the stimuli in the task. For example, such deterministic noise could be driven by differences in where people look, or for how long they look, or by whether they were fidgeting or scratching their nose \citep{Musall2019}. In addition to this conceptual limitation in measuring deterministic noise, parameter recovery simulations suggest that our estimation method also slightly underestimates deterministic noise (see Figure \ref{fig:paramrecover_main}, Supplementary Figure S5, S6). As a result, from both a conceptual and methodological perspective, it is possible that the remaining 86\% of the decision noise that is not stimulus-driven noise, could be deterministic.

@@

Such a reduction in the strength of reward coding in exploration, is consistent with our recent work using a drift diffusion model (DDM) to model explore-exploit decisions \citep{Feng21}. In the drift diffusion model, changes in behavioral variability can be driven by changes in the decision threshold (smaller threshold = more noise) or changes in the signal-to-noise ratio with which reward is encoded (lower SNR = more noise). By fitting both choices and response times, we were able to distinguish between these two accounts showing that the majority of the horizon-change in variability was driven by changes in SNR not threshold. However, this model could not determine whether the changes in SNR were driven by signal or noise. By showing that the change in deterministic and random noise have approximately the same ratio, the present work suggests that this SNR change is driven by changes in reward-signal processing, not noise. Of course, to truly see whether changes in signal or noise are driving random exploration will require more direct measurements of neural processing such as with neuroimaging and electrophysiology \citep{Tomov20, Ebitz18,costa1,costa22}

@@

80 participants (ages 18-25, 37 male, 43 female) from the University of Arizona undergraduate subject pool participated in the experiment. 15 were excluded on the basis of performance, using the same exclusion criterion as in \cite{wilson2014}. In this exclusion criteria, we measured the accuracy of each participant's choices by calculating the percentage of times that a participant chose the bandit with the higher underlying mean payouts in the last choice of a long horizon game, intuitively people should figure out which bandit has a higher mean payout by the last trial and should have an accuracy measure significantly above 50\%, specifically, we computed the likelihood that the measured accuracy can be achieved by making a completely random choice between the two options and excluded participants with a likelihood greater than 0.1\%, in other words, participants who didn't show an accuracy significant above chance with $p < 0.001$ were excluded in the analysis. This left 65 for the main analysis. Note that including the 15 badly performing subjects did not change the main results (Supplementary Figures S1, S2, S11)

@@

Each game consisted of 5 or 10 choices. Every game started with a fixation cross, then a bar of boxes appeared indicating the horizon for that game. For the first 4 trials - the instructed `forced-choice' trials, we highlight the box on one of the bandits to instruct the participant to choose that option. On these trials, they have to press the corresponding key to reveal the outcome. From the fifth trial, boxes on both bandits will be highlighted and they are free to make their own decision. There was no time limit for decisions. During free choices participants could press either the left arrow key or right arrow key to indicate their choice of left or right bandit. The score feedback was presented for 300ms. The task was programmed using Psychtoolbox in MATLAB \citep{psychtoolbox1, psychtoolbox2}.

The first four trials of each game were forced-choice trials, in which only one of the options was available for the participant to choose. We used these forced-choice trials to manipulate the relative ambiguity of the two options, by providing the participant with different amounts of information about each bandit before their first free choice. The four forced-choice trials set up two uncertainty conditions: unequal uncertainty(or [1 3]) in which one option was forced to be played once and the other three times, and equal uncertainty(or [2 2]) in which each option was forced to be played twice. After the forced-choice trials, participants made either 1 or 6 free choices (two horizon conditions).

\subsection*{Model-based analysis}
We modeled behavior on the first free choice of the Horizon Task using a version of the logistic choice model in \cite{wilson2014} that was modified to differentiate deterministic noise from random noise. Because the stimuli are identical in the repeated games, by definition, deterministic noise remains the same in repeated games, whereas random noise can change.
